---
title: "Week 5 Homework"
author: "Madison Muhle"
date: 2026-05-10
format:
  pdf:
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
        \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
execute:
  message: false
  warning: false
---

# Set Up

## Packages

```{r}
#| label: packages

# general use and data cleaning
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrapping long axis labels
library(scales)

# reading in .xlsx files
library(readxl)

# visualizing missing data
library(naniar)

# arranging multiple plots into panels
library(patchwork)
```

## Data

```{r}
#| label: data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects")

# site information
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# water quality information
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+"))
```

## Class Cleaning Code

```{r}
#| label: ncos-site-object

# creating new object from sites with only quarterly NCOS sites
ncos_sites <- sites |>
  # keep only NCOS sites sampled four times a year
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```

```{r}
#| label: cleaning-taxon-list

# create cleaned taxon list with snake_case names for joining
taxon_list_clean <- taxon_list |>
  # convert verbatim names to snake_case so they match aq_ins column names
  mutate(verbatim_name = to_any_case(verbatim_name, case = "snake"))
```

```{r}
#| label: water-quality-cleaning

# create summarized water quality object with median values per site-date
water_quality_clean <- water_quality |>
  # clean column names
  clean_names() |>
  # filtering join: keep only rows from quarterly NCOS sites
  semi_join(ncos_sites, by = "site") |>
  # combine date and site into one column for joining later
  unite("date_site", date, site, remove = TRUE) |>
  # group by each unique site-date combination
  group_by(date_site) |>
  # calculate median pH, salinity, and dissolved oxygen per site-date
  summarize(med_ph  = median(p_h,                  na.rm = TRUE),
            med_sal = median(salinity_ppt,          na.rm = TRUE),
            med_do  = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |>
  # ungroup after summarizing
  ungroup() |>
  # remove known salinity outlier at this site-date
  filter(date_site != "2022-08-12_NVBR")
```

```{r}
#| label: cleaning-aq-ins

# create long-format clean data frame with abundance, water quality, and taxonomy
aq_ins_clean <- aq_ins |>
  # clean column names
  clean_names() |>
  # filtering join: keep only rows from quarterly NCOS sites
  semi_join(ncos_sites, by = c("site")) |>
  # rename date column for clarity
  rename(date = date_on_vial) |>
  # select only columns needed for analysis
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae,
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |>
  # keep only filtered beaker samples
  filter(sample_type %in% c("FB 250", "FB250")) |>
  # pivot to long format so each row is one taxon observation
  pivot_longer(cols      = ostracod:annelida_polychaete,
               names_to  = "taxon",
               values_to = "abundance") |>
  # group by date, site, and taxon to summarize across replicates
  group_by(date, site, taxon) |>
  # calculate average abundance per liter (total filtered through 7.5 L)
  summarize(ave_lit = sum(abundance, na.rm = TRUE) / 7.5) |>
  # ungroup after summarizing
  ungroup() |>
  # create date_site column for joining with water quality
  unite("date_site", date, site, remove = FALSE) |>
  # attach water quality data by site-date
  left_join(water_quality_clean, by = c("date_site")) |>
  # attach taxonomic information by taxon name
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |>
  # add full site names for readable axis labels
  mutate(site_full = case_when(
    site == "NVBR"  ~ "North of Venoco Bridge",
    site == "NEC"   ~ "South of Venoco Bridge",
    site == "NMC"   ~ "Main Channel",
    site == "NPB"   ~ "South of Phelps Bridge",
    site == "NPB1"  ~ "North of Phelps Bridge",
    site == "NPB2"  ~ "Phelps Road",
    site == "NWP"   ~ "West Pond",
    site == "NDC"   ~ "Devereux Creek"
  )) |>
  # order sites by median salinity (highest to lowest) for plot axis
  mutate(site_full = fct_reorder(site_full, med_sal,
                                 .fun = "median", .na_rm = TRUE),
         site_full = fct_rev(site_full))
```

## Class Visualization Code

```{r}
#| label: visualizing-salinity
#| fig-width: 8
#| fig-height: 6

# salinity panel: individual survey measurements + median per site
salinity <- ggplot(data    = aq_ins_clean,
                   mapping = aes(x = site_full,
                                 y = med_sal)) +
  # individual salinity measurements at each survey
  geom_jitter(height = 0,
              width  = 0.1,
              shape  = 21) +
  # wrap long site names on the x-axis
  scale_x_discrete(labels = label_wrap(width = 15)) +
  # median salinity per site shown as red point
  stat_summary(fun   = median,
               color = "red",
               size  = 1) +
  # readable axis labels and title
  labs(x     = "Site",
       y     = "Median salinity (ppt)",
       title = "Site salinity") +
  # non-default theme
  theme_bw()
```

```{r}
#| label: copepod-cleaning

# filter to copepods and log + 1 transform abundance
copepods_df <- aq_ins_clean |>
  # keep only copepod rows
  filter(taxon == "copepod") |>
  # log + 1 transform to reduce skew and handle zeros
  mutate(log_ave_lit = log(ave_lit + 1))
```

```{r}
#| label: visualizing-copepods
#| fig-width: 10
#| fig-height: 6

# copepod panel: individual observations + median per site
copepods <- ggplot(data    = copepods_df,
                   mapping = aes(x = site_full,
                                 y = log_ave_lit)) +
  # individual log-transformed abundance values per survey
  geom_jitter(height = 0,
              shape  = 21,
              width  = 0.1) +
  # median abundance per site shown as blue point
  stat_summary(geom  = "point",
               fun   = "median",
               size  = 4,
               color = "blue") +
  # wrap long site names on the x-axis
  scale_x_discrete(labels = label_wrap(15)) +
  # readable axis labels and title
  labs(x     = "Site",
       y     = "log + 1 transformed average abundance/L",
       title = "Copepods") +
  # non-default theme
  theme_bw()
```
---

# Problem 1: Visualizing differences across sites in major taxon abundance

## 1a. Plan your figure

For the ostracod panel in 1c:

# x-axis: `site_full` — sites ordered by median salinity (highest to lowest)
# y-axis: `log_ave_lit` — log + 1 transformed average ostracod abundance per liter
# geometry to represent average abundance/L: `geom_jitter()` — one open circle per site × date survey, with slight horizontal jitter to reduce overplotting
# geometry to represent medians: `stat_summary(fun = "median")` — a larger filled point in a contrasting color and larger size, overlaid on the jittered observations

## 1b. Wrangle data

```{r}
#| label: ostracod-cleaning

# filter to ostracods and log + 1 transform abundance, mirroring copepod wrangling
ostracods_df <- aq_ins_clean |>
  # keep only ostracod rows
  filter(taxon == "ostracod") |>
  # log + 1 transform to reduce right skew and handle zero counts
  mutate(log_ave_lit = log(ave_lit + 1))

# display 10 random rows to verify wrangling output
slice_sample(ostracods_df, n = 10)
```

## 1c. Code the ostracod figure

```{r}
#| label: visualizing-ostracods
#| fig-width: 10
#| fig-height: 6

# ostracod panel: individual observations + median per site
ostracods <- ggplot(data    = ostracods_df,
                    mapping = aes(x = site_full,
                                  y = log_ave_lit)) +
  # individual log-transformed abundance values per survey as open circles
  geom_jitter(height = 0,
              shape  = 21,
              width  = 0.1) +
  # median abundance per site: larger orange diamond, visually distinct from
  # the small open jitter circles by color, size, and shape type
  stat_summary(geom  = "point",
               fun   = "median",
               size  = 4,
               color = "orange",
               shape = 18) +
  # wrap long site names to prevent overlap on x-axis
  scale_x_discrete(labels = label_wrap(15)) +
  # readable axis labels and title naming the taxonomic group
  labs(x     = "Site",
       y     = "log + 1 transformed average abundance/L",
       title = "Ostracods") +
  # non-default clean theme
  theme_bw()

# display ostracod plot
ostracods
```

## 1d. Multipanel figure

```{r}
#| label: multipanel-figure
#| fig-width: 8
#| fig-height: 14

# stack all three panels vertically using patchwork:
# salinity on top, copepods in middle, ostracods on bottom
salinity / copepods / ostracods
```

## 1e. Interpretation

Median copepod log-abundance (shown by the blue points) is higher at West Pond and Devereux Creek, the lower-salinity sites on the left side of the x-axis, than at higher-salinity sites like North of Venoco Bridge, where the blue median point is near zero. Ostracods show a similar pattern: the orange median diamond is highest at West Pond and decreases toward  higher-salinity sites, suggesting that ostracod abundance has lower salinity conditions.

---

# Problem 2: Visualizing total abundance as a function of salinity

## 2a. Plan the figure

# x-axis: `med_sal` — median salinity (ppt) recorded on the sampling date
# y-axis: `log_ave_lit` — log + 1 transformed average abundance per liter
# color: `taxon_display` — taxon name in title case (one distinct color per taxon)
# geometry: `geom_point()` — scatter of individual site × date observations
# facets: `facet_wrap(~ taxon_display, scales = "free")` — one panel per taxon with free x- and y-axes 

